---
title: "Week 0: Getting Started with R"
subtitle: "Your first reproducible Quarto document"
format:
  html:
    embed-resources: true
editor: source
execute:
  echo: true
  warning: true
  message: false
---

## About this module

This module introduces R, RStudio, Markdown, and Quarto. You do not need previous programming or statistics experience. Work through the sections in order and make changes directly in this file.

The module is complete when you can restart R and render this document successfully. You do not need to submit or share it.

## Understand the tools

| Tool | What it does |
|---|---|
| R | Performs calculations and runs R code |
| RStudio | Provides an interface for working with R |
| R package | Adds functions and capabilities to R |
| Quarto document | Combines Markdown text, code, output, and visualizations |

Install R before installing RStudio. If either program is missing, follow the instructions in the [Week 0 README](README.md) before continuing.

## Work in the source editor

This file is configured to use RStudio's source editor. You will see the Markdown syntax and R code that produce the finished document.

A Quarto document contains two main kinds of content:

- **Markdown** communicates headings, paragraphs, lists, emphasis, and links.
- **Code chunks** contain executable R code and produce output.

## Practice Markdown

Markdown headings begin with number signs:

```markdown
## Major section

### Smaller section
```

Other common syntax includes:

```markdown
This is **important** and this is *emphasized*.

- First item
- Second item

The `library()` function loads an installed package.

Visit the [R Project website](https://www.r-project.org/).
```

### Your turn

Replace the text below with two or three things you hope to learn with R. Format them as a Markdown list.

- Replace this item.
- Replace this item.

## Run your first R code

The gray area below is an R code chunk. Click inside it and use its green Run button. The result should appear beneath the chunk.

```{r}
2 + 2
```

The opening line tells Quarto that the chunk contains R code. The three backticks at the end close the chunk. Regular explanatory text belongs outside code chunks.

## Create and inspect objects

The assignment operator `<-` stores a value under a name.

```{r}
course_name <- "Getting Started with R"
number_of_sessions <- 10

course_name
number_of_sessions
```

Change `course_name` and `number_of_sessions`, then run the chunk again. Choose clear object names that describe what they contain.

## Perform a calculation

Objects can be used in later calculations.

```{r}
minutes_per_session <- 45

total_minutes <- number_of_sessions * minutes_per_session

total_minutes
```

Change `minutes_per_session` and observe how the result changes.

## Work with text

Replace the placeholder with your preferred name, then run the chunk.

```{r}
first_name <- "Replace this text"

message("Hello, ", first_name, "! R is working.")
```

## Install and load a package

Installing and loading are different operations:

- `install.packages()` puts a package on the computer. You generally do this once.
- `library()` makes an installed package available in the current R session. You generally do this after starting a new session.

The installation chunk is not evaluated when the full document renders. If the tidyverse is not installed, place the cursor inside the chunk and run it manually once.

```{r}
#| eval: false

install.packages("tidyverse")
```

Now load the package:

```{r}
library(tidyverse)
```

If R reports `there is no package called 'tidyverse'`, return to the installation chunk and run it.

## Create a small dataset

A tibble is a rectangular dataset with rows and columns. Edit the activities and minutes to make this example personally meaningful.

```{r}
practice_data <- tibble(
  activity = c("Reading", "Exercise", "Music"),
  minutes = c(30, 45, 20)
)

practice_data
```

After editing, make sure that the number of activities still matches the number of time values.

## Create a visualization

The following code uses ggplot2, which was loaded as part of the tidyverse. Run the chunk and inspect the visualization that appears below it.

```{r}
ggplot(
  practice_data,
  aes(x = activity, y = minutes)
) +
  geom_col(fill = "#006778") +
  labs(
    title = "How I might spend some free time",
    x = NULL,
    y = "Minutes"
  ) +
  theme_minimal()
```

Change the title so that it accurately describes your edited data.

### Describe the result

Replace this paragraph with one or two sentences describing something visible in your table or visualization.

## Inspect the computing environment

The following commands report the R version and operating system. This information can be useful when troubleshooting or reproducing an analysis.

```{r}
R.version.string
Sys.info()[["sysname"]]
```

You do not need to memorize or interpret all the technical details returned by `sessionInfo()`.

```{r}
sessionInfo()
```

## Render the complete document

Complete these steps:

1. Save this `.qmd` file.
2. In RStudio, restart R using **Session > Restart R**.
3. Select **Render**.
4. Confirm that the resulting HTML document contains your text, calculations, table, and visualization.

Rendering after a restart tests whether the document contains everything needed to reproduce its own output.

## Troubleshooting

### RStudio cannot find R

Confirm that R was installed before RStudio. Close RStudio, reinstall R from the official R website, and reopen RStudio.

### A package cannot be found

Run the installation chunk manually and then rerun `library(tidyverse)`.

### Code appears as text

Confirm that the code is between an opening line containing three backticks followed by `{r}` and a closing line containing three backticks.

### Markdown produces an error in R

Markdown text may have been placed inside an R code chunk. Move prose outside the chunk boundaries.

### The document works interactively but will not render

Restart R and run the document from the beginning. Look for an object that was created in the console but not created inside the document.

### Output appears in the console instead of below a chunk

Make sure you opened the `.qmd` file in RStudio and are running code from a code chunk. RStudio preferences and versions can affect where interactive output appears, but the rendered HTML should still contain the output.

## Completion checklist

- [ ] R is installed.
- [ ] RStudio opens successfully.
- [ ] I worked in the source editor.
- [ ] I edited a Markdown heading or list.
- [ ] I distinguished Markdown from R code chunks.
- [ ] I ran code and inspected its output.
- [ ] I installed and loaded the tidyverse.
- [ ] I created a small tibble.
- [ ] I created a ggplot visualization.
- [ ] I saved the `.qmd` file.
- [ ] I restarted R.
- [ ] I rendered the complete document successfully.

