---
title: "Week 1: Import data, manipulate it, then visualize it"
format:
  html:
    embed-resources: true
editor: source
execute:
  echo: true
  warning: true
  message: false
title-block-banner: true
toc: true
---

Which teams won the most games in the 2025 NFL regular season?

**Data credit:** The CSV uses game and schedule data maintained by Lee Sharpe and retrieved with [`nflreadr::load_schedules(2025)`](https://nflreadr.nflverse.com/reference/load_schedules.html).

## Load Packages

Only run this chunk if you haven't installed tidyverse on your computer before
```{r}
#| eval: false
install.packages("tidyverse")
```

But always run this chunk
```{r}
library(tidyverse)
```

## Import data

Use the read_csv function to import the data located "data/games_2025.csv".

Store that data in a variable called 'games'

Print the variable 'games' to the screen
```{r}
games <- read_csv("data/games_2025.csv", show_col_types = FALSE)

games
```
Open the help panel and search for read_csv.
What other types of data can be imported?

In this dataset, each row is a single game from the 2025 season.

Which columns will we use to count how many wins each team had?

Here are some useful functions to help us inspect data.

How many rows are in the dataset?
```{r}
nrow(games)
```
What are the names of the variables in the dataset?
```{r}
colnames(games)
```
How many games were played this season?
```{r}
count(games)
```

How many games were played in Jacksonville?
```{r}
count(games, home_team == "JAX")
```
Wait... weren't some games played in London?

Something to look into later.
Apparently home team and location of game are different.

## Tidy data

To start, here is our dataset
```{r}
games
```

What does the select function do?

```{r}
games |>
  select(week, away_team, away_score, home_team, home_score) |>
  slice_head(n = 6)
```
What did the slice_head function do?

If you aren't sure, open the help panel and search for slice.

Read the above code as:
1. start with games
2. THEN SELECT the following variables: week, away_team, away_score, home_team, home_score
3. THEN SLICE the first 6 rows (keep only the first six rows)


What does the count function do?
```{r}
games |>
  count(week)
```
What does the mutate function do?
```{r}
games |>
  mutate(score_difference = home_score - away_score) |>
  select(week, away_team, home_team, score_difference) |>
  slice_head(n = 6)
```

Read the code above as:
1. start with games
2. THEN MUTATE (in this case, create) a new variable called 'score_difference' by subtracting away_score from home_score
3. THEN SELECT the following variables (week, away_team, home_team, score_difference)
4. THEN SLICE the first 6 rows

## Count team wins

The data have one row per game.
Your chart needs one row per team.

So what will you do?

There are a million strategies you could use. I've written one below.

1. start with games
2. THEN MUTATE a new variable called 'score_difference' by subtracting away_score from home_score
3. THEN MUTATE a new variable called 'winner' by recording home_team if score_difference > 0, otherwise recording away_team
    * mutate(newvariable = ifelse(score_difference>0, home_team, away_team))
    * You can remember how to use the ifelse function with the acronym WTF
    * WHAT: score_difference>0
    * TRUE: if true, record the home_team as the winner
    * FALSE: if false, record the away_team as the winner
4. THEN COUNT the winners
5. THEN store the output in a variable called num_team_wins


Problem with this algorithm: What if there is a tie?

Advanced mode: Fix the algorithm to handle ties.


```{r}
#| eval: false

games |>
  mutate(score_difference = home_score - away_score) |>
```
## Make the chart

```{r}
#| eval: false
#| fig-width: 10
#| fig-height: 10

num_team_wins |>
  ggplot(aes(x = wins, y = team)) +
  geom_col()+
  labs(
    title = "",
    x = "Regular-season wins",
    y = NULL,
    caption = "hmm...."
  ) +
  theme_ +
  theme()
```

Can you change the colors?

Can you change the fonts?

Can you change the size of the fonts?

Can you flip the chart?
