---
title: "Week 2: Manipulate data with dplyr"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: true
  message: false
---

Last week we learned which teams won the most.

This week we'll find out which teams won (or lost) by the most.

And, along the way, learn a few more dplyr tricks and tips.

## Load packages

```{r}
library(tidyverse) # You should already have installed this package in previous weeks
source("helper-functions.R") # This is a place where you will paste useful functions to reuse
```

## Import data

```{r}
games <- read_csv("data/games_2025.csv", show_col_types = FALSE)
glimpse(games)
```

## Tidy data

1. START with games
2. THEN SELECT the following columns: season, home_team, away_team, home_score, and away_score
3. STORE (->) the new dataset in 'season_games'

```{r}
games |>
  select(season, home_team, away_team, home_score, away_score) -> season_games
```

1. START with season_games
2. THEN FILTER out any rows (games) that have an NA instead of value
3. STORE the new dataset in 'completed_games'

```{r}
season_games |>
  filter_out(is.na(home_score) | is.na(away_score)) -> completed_games
```

1. START with completed_games
2. THEN RENAME the following columns: home_team -> team, away_team -> opponent, home_score -> points_scored, away_score -> points_allowed
3. THEN MUTATE (create) a new variable called 'venue' with the value 'Home' in every row
4. STORE the new dataset in 'home_team_games'

```{r}
completed_games |>
  rename(team = home_team,
         opponent = away_team,
         points_scored = home_score,
         points_allowed = away_score) |>
  mutate(venue = "Home") -> home_team_games
```

Now do the same for away_team_games

1. START with completed_games
2.
3.
4.

```{r}
completed_games |>
  rename(team = away_team,
         opponent = home_team,
         points_scored = away_score,
         points_allowed = home_score)|>
  mutate(venue = "Away") -> away_team_games
```

1. START with home_team_games
2. THEN BIND_ROWS to append the away_team_games to bottom of the home_team_games (add new rows at the bottom of home_team_games)
3. MUTATE three new variables (columns)
    * scoring margin = points_scored - points_allowed
    * won = scoring_margin > 0
    * tied = scoring_margin == 0
4. STORE the new dataset in 'team_games'

```{r}
home_team_games |>
  bind_rows(away_team_games) |>
  mutate(scoring_margin = points_scored - points_allowed,
         won = scoring_margin > 0,
         tied = scoring_margin == 0) -> team_games
```

## Summarize team performance

1. START with team_games
2. THEN GROUP_BY team (imagine splitting the dataset into groups such that each group contains all the rows for 1 specific team. In this case, each group will have 17 rows: one for each game the team played)
3. THEN SUMMARIZE for each group (team, in this case), total games played, total wins, total ties, average points scored, average points allowed, and average scoring margin.
4. STORE the new dataset in team_summary

```{r}
team_games |>
  group_by(team) |>
  summarise(games_played = n(),
            wins = sum(won),
            ties = sum(tied),
            average_points_scored = mean(points_scored),
            average_points_allowed = mean(points_allowed),
            average_scoring_margin = mean(scoring_margin)) -> team_summary
```

1. START with team_summary
2. THEN ARRANGE (sort) the data by wins
3. THEN SLICE_HEAD the data so we only see the top (errrr.... bottom) 5

```{r}
team_summary |>
  arrange(wins, team) |>
  slice_head(n = 5)
```
1. START with team_summary
2. THEN ARRANGE (sort) the data by wins IN DESCENDING ORDER (top to bottom)
3. THEN SLICE_HEAD the data so we only see the top 5

```{r}
team_summary |>
  arrange(desc(wins), team) |>
  slice_head(n = 5)
```

1. START with team_summary
2. THEN ARRANGE (sort) the data by average points scored IN DESCENDING ORDER (top to bottom)
3. THEN SLICE_HEAD the data so we only see the top 6
```{r}

```

1. START with team_summary
2. THEN ARRANGE (sort) the data by average scoring margin IN DESCENDING ORDER (top to bottom)
3. THEN SLICE_HEAD the data so we only see the top 7

```{r}

```

## Create the weekly visual

1. START with team_summary
2. THEN pass it to ggplot
3. Set the x-axis to average scoring margin (try changing this to another column in team_summary)
4. Reorder (sort) the y-axis by team with the highest scoring margin
5. Add a vline to the plot (try changing the color or linetype)
6. Add points to the plot (try changing color, size, shape)
7. Add titles (labels) to the plot
8. Add a theme to the plot!
9. Keep customizing the theme

```{r}
#| fig-width: 9
#| fig-height: 8

team_summary |>
  ggplot(aes(x = average_scoring_margin, y = reorder(team, average_scoring_margin))) +
  geom_vline(xintercept = 0, color = "grey65", linetype = "dashed") +
  geom_point(size = 2.8, color = "#176B87") +
  labs(title = "Average scoring margin by NFL team",
       subtitle = "2025 regular season",
       x = "Average scoring margin (points per game)",
       y = NULL,
       caption = "Data: nflverse/nfldata via nflreadr; 2025 regular season") +
  theme_nfl_lab()
```
